Keep your Laziness in Check

### 2018-03-03 paper/main.tex
[paper] more work on section 2

Lazy evaluation gives tremendous power to functional programmers
~\cite{WhyFPMatters} and is actively used in various practical
applications. For instance, the big-data framework Spark uses laziness
to drive computations on demand\cn, Apple's programming language Swift
leverages it to initialize expensive resources as needed\cn, and many
functional reactive programming web frameworks crucially rely on
it\cn \leo{This point might need more exposition about frp}. Haskell
is the most popular programming language that allows programmers to
enjoy the power of lazy evaluation by default. However, that power
does not come without a price.
\leo{Didn't we have an FRP quote by someone that Laziness was the third source of
bugs or something? Work that in?}
\hzh{Ryan Trinkle (author of Reflex) said laziness was the 3rd most difficult thing to
get right, but let's check with him on whether we can quote him on that.}

@@

How, then, can we test and catch these laziness bugs during
development? Is it possible to precisely specify laziness behavior
just as we would for functional correctness?
%
In this paper, we demonstrate techniques for observing laziness in
Haskell, and propose a system of creating specifications of lazy
Haskell programs. We implement these ideas as a Haskell library. For
concreteness, consider the following small example.

\leo{Insert take example here. There is enough space to have both code
and contributions in this page. yay :) }\caleb{YES!}

In the next section we work through two more examples, functional
queues and functional maps, demonstrating both the difficulties of
programming with laziness and how our approach can be used to
alleviate them.
%
In the rest of the paper, we present our technical contributions:
\begin{itemize}
\item We introduce a novel approach to specifying the laziness behavior
of functions that is {\em precise} and {\em compositional}. Such
specifications are expressive enough to capture exactly how much
evaluation occurs within a data structure
%\leo{Also add
%drawbacks/partiality here? If not here, where in the intro?} \hzh{I'm
%not sure if we should bring in partiality (in the sense of
%divergence/nontermination) yet since that seems like a rather
%technical detail of the spec}
\item We propose a framework for testing against such specifications.\caleb{Automated testing? Can be more specific}
\item We improve upon the random function generators from QuickCheck\cn to generate functions with arbitrary laziness behaviors.
\item We demonstrate non-trivial generic programming techniques to apply StrictCheck to almost all data types.\caleb{I don't know
the meaning of non-trivial and almost-all}
\end{itemize}
%
Section~\ref{sec:related} discusses related work. We conclude and draw
directions for future work in Section~\ref{sec:concl}.

@@

To describe the laziness behavior, we need to first introduce a few
types that describe how lists are used. We use the
term \textit{demand} to refer to the shape of such usage. Concretely,
a \textit{demand} is a shape that describes the used part of a data
structure. For lists, \textit{demands} will be prefixes of lists. We
use the type \lk{ListD} to represent such demands (\Cref{fig:ListD}).
\leo{Why is this a figure and not just inline? Especially if we're inlining
the figure!}
%
\begin{figure}[H]
\centering
\begin{inlinecode}
  data ListD a = NilD
               | ConsD (Thunk a) (Thunk (ListD a))
\end{inlinecode}
\caption{\lk{ListD} type}
\label{fig:ListD}
\end{figure}
%
The \lk{Thunk} type used in \lk{ListD} has 2 constructors: \lk{E}
for \textit{evaluated} and \lk{T}, a nullary constructor,
for \textit{thunk}. In \lk{ListD}, \lk{Thunk} wraps around all the
list constructor fields, and this allows us to represent lack of
further evaluation by using the \lk{T} constructor. For example,
given the list \lk{[1, 2, 3]}, the value \lk{ConsD T (ConsD T T)}
represents the \textit{demand} that reveals the first 2 \lk{Cons}
constructors of the list, but does not evaluate any other parts of the
list. To improve readability of demand values, we will also use the
syntax \lk{\_ : \_ : \_} where appropriate.

In Haskell, to trigger the evaluation of a lazy function, we need to
exert some \textit{demand} on the result of that function. This
suggests a natural flow for the specification of a lazy function ---
given the demand on the output, the specification should calculate the
demands on the input. However, this is not enough in general because
the demand on the input could depend on the value of the inputs as
well. The \lk{take} function introduced in \Cref{sec:intro} is such a
case since the length of the output from \lk{take} depends on its
integer argument.

@@

For the second example, we turn to a different functional data structure where
both a strict and a lazy variant can be situationally useful: maps. There are
two reasonable laziness behaviors that implement a map interface; we could
implement a completely strict map, or we could lazily evaluate the values stored
as needed.
\footnote{A third option could be to have lazy keys, but they would
necessarily be forced at every insertion or lookup, with no gain.}
Haskell's standard library provides these two variants in the form of
\lk{Data.Map.Strict} and \lk{Data.Map.Lazy}. However, there is no way
to {\em specify} the difference between the two versions! At best, one
could hope for a hand-crafted suite of unit tests containing bottoms
to observe a difference in termination.
\leo{One thing we could do is ignore the knapsack and expand on the above.}

Moreover, it is important to be able to distinguish whether such a datatype
follows a desired laziness behavior. For instance, the standard functional
solution for dynamic programming problems crucially uses laziness \hzh{is this
solution really standard? Is there some work we can cite to establish this
approach as the canonical approach for dynamic programming in functional
programming?}. In particular, consider the famous knapsack problem.
%
The inputs of the problem are two sequences of numbers $w_i$ and $v_i$
where $i = 1, \dots, n$, and each of $w_i$ and $v_i$ specify the
weight and the value of the $i$th item. The solution presents the
maximum value achievable using a subset of the items such that their
combined weight is under a limit $W$.
%
The usual dynamic programming approach for knapsack maintains a
table $OPT[][]$, where the entry $OPT[j][k]$ keeps the best value
achievable using only items $1\dots j$ and a weight limit of $k$. The
table satisfies the following recurrence relation: $$OPT[j][k]
= \max(OPT[j-1][w], w_j + OPT[j-1][k-w_j])$$

To efficiently compute this table using imperative langauges, we look
for shared sub-problems of the original problem, and design a path
along the solution table that maximizes sharing while filling out the
answers.
%
However, lazy evaluation automatically finds this path if we use lazy
maps to represent the solution table according to the recurrence
relation \hzh{We can probably make this code prettier}
\begin{inlinecode}
ws, vs :: [Int]
n,  w  :: Int

solutionPre :: Map (Int, Int) Int -> Map (Int, Int) Int
solutionPre soln = fromList [ ((j, k),
                               if (j-1) < 0 || k - ws !! j < 0
                               then if ws !! j <= k then vs !! j else 0
                               else max (soln ! (j-1, k))
                                        (soln ! (j-1, k - ws !! j) + vs !! j)
                              )
                            | j <- [0..2], k <- [0..50] ]

solution :: Map (Int, Int) Int
solution = fix solutionPre
\end{inlinecode}

To query the optimal achievable values from the solution table, we can
just index into the map. When an entry from the map is looked up at
the first time, a sequence of evaluations is triggered, and their
results memoized in the lazy map automatically.

However, using the lazy variant of \lk{Map} here is critical for the
correctness of \lk{solution}. If we used a map strict in its values
instead, then \lk{fromList} would have eagerly evaluated the \lk{if}
expression. The \lk{if} expression contains value-level recursion to
the solution map itself, and this would manifest as an infinite loop
at runtime. This is an incredibly subtle bug related to laziness.

To catch this behavior through testing, we can write another specification that
captures this correctness condition. The well-foundedness \hzh{is this a word?}
of \lk{solution} depends on what is evaluated when the value at key \lk{(j, k)}
of the map is demanded. Let $\lk{solutionPre}_n$ be the $n$-times iterated
pre-fixpoint \lk{solutionPre . solutionPre . ... . solutionPre}, then if we
demand the value at \lk{(j, k)} from the result of $\lk{solutionPre}_n$, its
input map should not be evaluated at the \lk{(j, k)} entry since otherwise
evaluating the value at \lk{(j, k)} becomes an infinite loop at the fixpoint.

We can write another executable specification for this property
of $\lk{solutionPre}_n$ (\Cref{fig:spec-map}). Testing the implementation based on
strict maps also immediately produces a counter example.

\begin{figure}[H]
\centering
\hzh{TODO: specification for map}
\begin{inlinecode}
\end{inlinecode}
\caption{Specification for \lk{solutionPre}}
\label{fig:spec-map}
\end{figure}

>>>>

Lazy evaluation gives tremendous  to functional programmers
~\cite{WhyFPMatters} and is actively used in various practical
applications. For instance, the big-data framework Spark uses laziness
to drive computations on demand\cn, Apple's programming language Swift
leverages it to initialize expensive resources as needed\cn, and many
functional reactive programming web frameworks crucially rely on
it\cn \leo{This point might need more exposition about frp}. Haskell
is the most popular programming language that allows programmers to
enjoy the power of lazy evaluation by default. However, that power
does not come without a price.
\leo{Didn't we have an FRP quote by someone that Laziness was the third source of
bugs or something? Work that in?}
\hzh{Ryan Trinkle (author of Reflex) said laziness was the 3rd most difficult thing to
get right, but let's check with him on whether we can quote him on that.}

@@

How, then, can we test and catch these laziness bugs during
development? Is it possible to precisely specify laziness behavior
just as we would for functional correctness?
%
In this paper, we demonstrate techniques for observing laziness in
Haskell, and propose a system of creating specifications of lazy
Haskell programs. We implement these ideas as a Haskell library. For
concreteness, consider the following small example on the
function \lk{take}, which accepts an integer $n$ and a list, and returns
a prefix of the list with length $n$. If $n$ is larger than the length
of the original list, then the entire list is returned.
%
\begin{inlinecode}
take_spec :: Shaped a => Spec '[Int, [a]] [a]
take_spec =
  Spec $ \predict d n xs ->
    predict (nf n) (if n > length xs then E d else endThunk d)
  where
    endThunk :: PosDemand [a] -> Demand [a]
    endThunk = \case
      Nil'        -> T
      dx :% T     -> E (dx :% T)
      dx :% E dxs -> E (dx :% endThunk dxs)
\end{inlinecode}
%
\hzh{TODO: describe what this spec means in English}

In the next section we work through two more examples, functional
queues and functional maps, demonstrating both the difficulties of
programming with laziness and how our approach can be used to
alleviate them.
%
In the rest of the paper, we present our technical contributions:
\begin{itemize}
\item We introduce a novel approach to specifying the laziness behavior
of functions that is {\em precise} and {\em compositional}. Such
specifications are expressive enough to capture exactly how much
evaluation occurs within a data structure
%\leo{Also add
%drawbacks/partiality here? If not here, where in the intro?} \hzh{I'm
%not sure if we should bring in partiality (in the sense of
%divergence/nontermination) yet since that seems like a rather
%technical detail of the spec}
\item We propose a framework for automated testing against such specifications.\caleb{Automated testing? Can be more specific}
\item We improve upon the random function generators from QuickCheck\cn to generate functions with arbitrary laziness behaviors.
\item We discuss our generic programming-based impementation to apply StrictCheck to all algebraic data types.\caleb{I don't know
the meaning of non-trivial and almost-all}
\end{itemize}
%
Section~\ref{sec:related} discusses related work. We conclude and draw
directions for future work in Section~\ref{sec:concl}.

@@

To describe the laziness behavior, we need to first introduce a few
types that describe how lists are used. We use the
term \textit{demand} to refer to the shape of such usage. Concretely,
a \textit{demand} is a shape that describes the used part of a data
structure. For lists, \textit{demands} will be prefixes of lists. We
use the type \lk{ListD} to represent such demands.
%
\begin{inlinecode}
  data ListD a = NilD
               | ConsD (Thunk a) (Thunk (ListD a))
\end{inlinecode}
%
The \lk{Thunk} type used in \lk{ListD} has 2 constructors: \lk{E}
for \textit{evaluated} and \lk{T}, a nullary constructor,
for \textit{thunk}. In \lk{ListD}, \lk{Thunk} wraps around all the
list constructor fields, and this allows us to represent lack of
further evaluation by using the \lk{T} constructor. For example, given
the list \lk{[1, 2, 3]}, the value \lk{ConsD T (ConsD T T)} represents
the \textit{demand} that reveals the first 2 \lk{Cons} constructors of
the list, but does not evaluate any other parts of the list. To
improve readability of demand values, we will also use the
syntax \lk{\_ : \_ : \_} where appropriate, the underscore $\lk{\_}$
represents a thunk.

In Haskell, to trigger the evaluation of a lazy function, we need to
exert some demand on the result of that function. This suggests a
natural flow for the specification of a lazy function --- given the
demand on the output, the specification should calculate the demands
on the input. However, this is not enough in general because the
demand on the input could depend on the \textit{value} of the inputs
as well. The \lk{take} function introduced in \Cref{sec:intro} is such
a case since the length of the output from \lk{take} depends on its
integer argument.

@@

For the second example, we turn to a different functional data structure where
both a strict and a lazy variant can be situationally useful: maps. There are
two reasonable laziness behaviors that implement a map interface; we could
implement a completely strict map, or we could lazily evaluate the values stored
as needed.
\footnote{A third option could be to have lazy keys, but they would
necessarily be forced at every insertion or lookup, with no gain.}
Haskell's standard library provides these two variants in the form of
\lk{Data.Map.Strict} and \lk{Data.Map.Lazy}. However, there is no way
to {\em specify} the difference between the two versions! At best, one
could hope for a hand-crafted suite of unit tests containing divergent
terms to observe a difference in termination.
\leo{One thing we could do is ignore the knapsack and expand on the above.}

Moreover, it is important to be able to distinguish whether such a datatype
follows a desired laziness behavior. For instance, the standard functional
solution for dynamic programming problems crucially uses laziness \hzh{is this
solution really standard? Is there some work we can cite to establish this
approach as the canonical approach for dynamic programming in functional
programming?}. Let us consider the famous knapsack problem.
%
The inputs of the problem are two sequences of numbers $w_i$ and $v_i$
where $i = 1, \dots, n$, and each of $w_i$ and $v_i$ specify the
weight and the value of the $i$th item. The solution presents the
maximum value achievable using a subset of the items such that their
combined weight is under a limit $W$.
%
The usual dynamic programming approach for knapsack maintains a
table $OPT[][]$, where the entry $OPT[j][k]$ keeps the best value
achievable using only items $1\dots j$ and a weight limit of $k$. The
table satisfies the following recurrence relation: $$OPT[j][k]
= \max(OPT[j-1][w], w_j + OPT[j-1][k-w_j])$$

To efficiently compute this table using imperative langauges, we look
for shared sub-problems of the original problem, and design a path
along the solution table that maximizes sharing while filling out the
answers.
%
However, in Haskell, we can define a function that computes one step
of the recurrence relation, and take the fixpoint of this function as
the final solution table. Laziness allows us to define such a solution
table without actually computing all of its entries. When an entry is
demanded, laziness again helps us by only evaluating the required
entries, freeing us from having to design an optimal path along the
solution table.
%
\begin{inlinecode}
weights :: [Int]
values  :: [Int]
limit   :: Int

solutionStep :: Map (Int, Int) Int -> Map (Int, Int) Int
solutionStep soln =
  fromList [((j, k), knapsack j k) |
            j <- [0 .. length weights-1], k <- [0 .. limit]]
  where
    knapsack j k = if j - 1 < 0 || k - weights !! j < 0
                   then if weights !! j <= k then values !! j else 0
                   else max (soln ! (j-1, k))
                            (soln ! (j-1, k - weights !! j) + values !! j)

solution :: Map (Int, Int) Int
solution = fix solutionStep

\end{inlinecode}
%
To query the optimal achievable values from the solution table, we can
just index into the map. When an entry from the map is looked up at
the first time, a sequence of evaluations is triggered, and their
results memoized in the lazy map automatically.

Using the lazy variant of \lk{Map} here is critical for the
correctness of \lk{solution}. If we used a map strict in its values
instead, then \lk{fromList} would have eagerly evaluated the \lk{if}
expression. The \lk{if} expression contains value-level recursion to
\lk{solution} itself, and this would manifest as an infinite loop
at runtime. This is an incredibly subtle bug related to laziness.

To catch this behavior through testing, we can write another specification that
captures this correctness condition. The well-foundedness \hzh{is this a word?}
of \lk{solution} depends on what is evaluated when the value at key \lk{(j, k)}
of the map is demanded. Let $\lk{solutionStep}_n$ be the $n$-times iterated
step function \lk{solutionStep . solutionStep . ... . solutionStep}, then if we
demand the value at \lk{(j, k)} from the result of $\lk{solutionStep}_n$, its
input map should not be evaluated at the \lk{(j, k)} entry since otherwise
evaluating the value at \lk{(j, k)} becomes an infinite loop at the fixpoint.

We can write another executable specification for this property
of $\lk{solutionStep}_n$. Testing the implementation based on
strict maps also immediately produces a counter example.
%
\begin{inlinecode}
TODO: expectation-spec for solutionStep
\end{inlinecode}
%
